Embedding Intuitionistic into Classical Logic

### 2023-03-18 LPAR/LPAR-EIICL.tex
small fixes

\begin{definition}
	A $\Sigma$-\emph{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$, called the Kripke \emph{frame}, and an indexed family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$ we have $M_u\subseteq M_w$, $f^{I_w}|_{M_u} = f^{I_u}$ and $R^{I_w}|_{M_u} = R^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$}
	The model relation between $\mathcal M_u$, $v$ (mapping into $M_u$) and formulae $\varphi$ is defined as before, except in the cases of implication and $\forall$-quantification, where we set
	\begin{itemize}
		\item $\mathcal M_u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $\mathcal M_w, v\not\models\varphi$ or $\mathcal M_w, v\models\psi$.
		\item $\mathcal M_u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $\mathcal M_w, v[m/a]\models\varphi[a/x]$.
	\end{itemize}
	We write $u, v\models\varphi$ if $\mathcal M_u, v\models \varphi$. $\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$. $\varphi$ is valid if it is satisfied by every Kripke structure.
	We denote the set of valid formulae with \emph{IQC} (Intuitionistic Quantified Calculus).
\end{definition}
In addition to propositional tautologies there are now quantified formulae which are classically valid but not intuitionistically, e.g., $\neg\forall x A(x)\to \exists x \neg A(x)$.

@@

\begin{definition}
	Let $\varphi$ be a propositional formula.\todo{genau genommen müsstest du noch sagen, dass $\nu$ eine freie Variable ist. Im letzten Fall der Definition benützen wir dann aber w gleichzeitig als freie und gebundene Variable}
 Define $\varphi^{\nu}$ inductively:
	\vspace*{-.1cm}
	\begin{itemize}
		\item $\bot^\nu = \bot$.
		\item $A^{\nu} = A(\nu)$ for every propositional variable $A$.
		\item $(\varphi\circ\psi)^\nu = \varphi^\nu\circ\psi^\nu$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^\nu = \forall w(\nu\preceq w\to\varphi^{w}\to\psi^{w})$ where $w$ is some new bound variable.
	\end{itemize}
	Let $K(\varphi)$ encode the theory of Kripke structures, i.e.
	\vspace*{-.1cm}
	\[K(\varphi) = \text{\normalfont PartialOrder}(\preceq)\wedge\forall u\forall w(u\preceq w\to \text{\normalfont Persistent}(u, w))\]
	\vspace*{-.4cm}
	with e.g.
	\vspace*{-.2cm}
	\begin{align*}
		\text{\normalfont PartialOrder}(\preceq) =&\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{\normalfont Persistent}(u, w)=&\:\bigwedge \{A(u)\to A(w)\:|\: \text{ $A$ occurs in $\varphi$}\}
	\end{align*}
	Then define
	$\varphi^{C} = K(\varphi)\to \varphi^{b}$
	where $b$ is a new free variable.
\end{definition}

@@

This is somewhat unsatisfying as we end up in a much more complex logic. However using the normal form from Lemma~\ref{lemma:propositional-normal-form}, i.e. assuming that $\varphi$ is of the form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$ for some flat clauses $\mathcal R$ and implicational clauses $\mathcal X$, further simplification is possible:
\begin{itemize}
	\item For flat clauses $(\bigwedge A_i\to\bigvee B_j)^\nu$ is
	$$\forall w\left(\nu\preceq w\to\bigwedge_i A_i(w)\to\bigvee_jB_j(w)\right).$$
	\item For implication clauses $((A\to B)\to C)^\nu$ is
	$$\forall w(\nu\preceq w\to(\forall k(w\preceq k\to A(k)\to B(k)))\to C(w))$$
\end{itemize}

@@

\begin{lemma}
	Suppose $\mathcal M = (M, T)$ is a counter-model to $\varphi^{CH}$, then so is $\mathcal M_T = (M_T, I_T)$.
\end{lemma}
\begin{proof}
	Let $v$ be a variable assignment. It follows directly from the definitions that $\mathcal M_T, v\not\models P(s)$ and $\mathcal M_T, v\models K(\varphi)$. It is easy to show that $\mathcal M_T, v\models \psi$ for $\psi\in\mathcal R^{bS}$ from persistency. Let $\psi\in\mathcal X^{bS}$, we show $\mathcal M_T, v[\psi_1\dots\psi_n/w] \models s\preceq w\to (w\preceq f_\psi(w)\to A(f_\psi(w)\to B(f_\psi(w)))\to C(w))$ for all $\psi_1\dots\psi_n\in M_T$. Let $\psi_1\dots\psi_n\in M_T$ and $m = g_{\psi_n}(\dots(g_{\psi_1}(s^I))$. If $\mathcal M, v[m/w]\models C(w)$ we are done. Otherwise $\mathcal M, v[m/w]\not\models s\preceq w$ or $\mathcal M, v[m/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w)) \to B(f_\psi(w))$. The former is impossible since $g_\psi$ is increasing for all $\psi$. For the latter to hold we must have $\mathcal M, v[m/w]\models w\preceq f_\psi(w)$, $\mathcal M, v[m/w]\models A(f_\psi(w))$ and $\mathcal M, v[m/w]\not\models B(f_\psi(w))$. If $\psi$ is fulfilled at $m$ we have $g_\psi(m) = m$, otherwise $g_\psi(m) = f_{[\psi]}^I(m)$. In any case $\mathcal M_T, v[\psi_1\dots\psi_n/w]\models A(f_\psi(w))$ and $\mathcal M_T, v[\psi_1\dots\psi_n/w]\not\models B(f_\psi(w))$. It follows that $\mathcal M_T, v[\psi_1\dots\psi_n/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w))$ and therefore $\mathcal M_T, v\models \psi$.
\end{proof}

@@

\begin{corollary}
	If a formula in normal form as in Lemma~\ref{lemma:propositional-normal-form} is intuitionistically invalid then there exists a Kripke counter-model that is a rooted tree of height and degree bounded by $|\mathcal X|$.
\end{corollary}

@@

It is known that a polynomial time translation between QBF and IPC must exist, since both problems are PSPACE complete~\cite{garey1979computers,statman1979intuitionistic}.
\todo{Wollen wir noch etwas sagen wie ``However the previous PSPACE result does not support the transfer of countermodels that we develop in this section''?}
For lack of space, we do not give the semantics of QBF in this paper and instead refer the reader to standard sources e.g.~\cite{series/faia/2009-185}.
The translation from the last section will serve as starting point for this section.

@@

\begin{definition}\label{def:fo-translation}
	Let $\varphi$ be a predicate formula. Define $\varphi^{\nu}$ inductively:
\todo{genau genommen müsstest du noch sagen, dass $\nu$ eine freie Variable ist. Im letzten Fall der Definition benützen wir dann aber w gleichzeitig als freie und gebundene Variable}
	\begin{itemize}
		\item $\bot^\nu = \bot$.
		\item $A(\vec t)^{\nu} = A(\vec t, \nu)$ for every predicate symbol $A$.
		\item $(\varphi\circ\psi)^\nu = \varphi^\nu\circ\psi^u$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^\nu = \forall w(\nu\preceq w\to\varphi^{w}\to\psi^{w})$.
		\item $(\exists x\varphi)^\nu = \exists x(E(x,\nu)\wedge\varphi^\nu)$.
		\item $(\forall x\varphi)^\nu = \forall w(i\preceq w\to \forall x(E(x, w)\to \varphi^w))$ where $w$ is a new bound variable.
	\end{itemize}
	where $w$ is some new bound variable. Let $K(\varphi)$ encode the theory of Kripke structures as well as the predicate $E$, i.e.
	\begin{align*}
		K(\varphi) = \:& \text{PartialOrder}(\preceq) \wedge \forall u \forall w (u\preceq w\to \text{DomainSubset}(u, w)) \wedge\\
		& \forall u(\text{World(u)}\to \text{DomainClosed}(u))\wedge \forall u\forall w (u\preceq w\to \text{Persistent}(u, w)).
	\end{align*}
	with e.g.:\vspace*{-.76cm}
	\begin{align*}
		\text{PartialOrder}(\preceq) = &\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{DomainSubset}(u, w) = &\:\forall x(E(x, u)\to E(x, w))\\
		\text{World}(u) = &\:\exists xE(x, u)\\
		\text{DomainClosed}(u) = &\:\bigwedge\left\{\forall\vec x(\vec E(\vec x, u)\to E(f(\vec x), u))\:|\:\text{$f$ a function in $\varphi$}\right\}\\
		\text{Persistent}(u, w) = &\:\bigwedge\left\{\forall\vec x(\vec A(\vec x, u)\to A(\vec x, w))\:|\:\text{$A$ a predicate in $\varphi$}\right\}
	\end{align*}
	Then define
	$\varphi^C = K(\varphi)\to\text{World}(b)\to \vec E(\vec a, b)\to \varphi^b$ where $\vec a$ contains all free variables occurring in $\varphi$ and $b$ is a new free variable.
\end{definition}

@@

\begin{proof}
	As in~\cite{otten2005clausal} we may assume that $\varphi$ is of the form $\psi\to P$. Recall that the transformation works by transforming $\psi$ into a set of equivalent assumptions $\left\lbrack \psi\right\rbrack$ of the desired form. We extend the given procedure with rules for quantifiers as follows:
	\begin{align*}
		\left\lbrack(\forall x \psi)\to A\right\rbrack&:= \left\lbrack\psi[s(\vec z)/x]\to A\right\rbrack\\
		\left\lbrack A\to (\forall x\psi)\right\rbrack&:= \{\forall x\chi[x/a]\:|\:\chi\in\left\lbrack A\to\psi[a/x]\right\rbrack\}\\
		\left\lbrack(\exists x\psi)\to A\right\rbrack&:= \{\forall x\chi[x/a]\:|\:\chi\in\left\lbrack A\to\psi[a/x]\right\rbrack\}\\
		A\to (\exists x\psi)&:= \left\lbrack A\to\psi[s(\vec z)/x]\right\rbrack
	\end{align*}
	where $s$ is a new function symbol, $a$ is a new free variable and $\vec z$ ranges over the free variables in $\psi$.
	While the generated set of clauses is in general not equivalent to $\psi$, section~\ref{section:herbrandiaztion} guarantees that it is at least equivalid.\hfill$\square$
\end{proof}

>>>>

\begin{definition}
	A $\Sigma$-\emph{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$, called the Kripke \emph{frame}, and an indexed family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$ we have $M_u\subseteq M_w$, $f^{I_w}|_{M_u} = f^{I_u}$ and $R^{I_w}|_{M_u} = R^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$}
	The model relation between $\mathcal M_u$, $v$ (mapping into $M_u$) and formulae $\varphi$ is defined as before, except in the cases of implication and $\forall$-quantification, where we set
	\begin{itemize}
		\item $\mathcal M_u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $\mathcal M_w, v\not\models\varphi$ or $\mathcal M_w, v\models\psi$.
		\item $\mathcal M_u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $\mathcal M_w, v[m/x]\models\varphi$.
	\end{itemize}
	We write $u, v\models\varphi$ if $\mathcal M_u, v\models \varphi$. $\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$. $\varphi$ is valid if it is satisfied by every Kripke structure.
	We denote the set of valid formulae with \emph{IQC} (Intuitionistic Quantified Calculus).
\end{definition}
In addition to propositional tautologies there are now quantified formulae which are classically valid but not intuitionistically, e.g., $\neg\forall x A(x)\to \exists x \neg A(x)$.

@@

\begin{definition}
	Let $\varphi$ be a propositional formula.
 Define $\varphi^{b}$ inductively:
	\vspace*{-.1cm}
	\begin{itemize}
		\item $\bot^b = \bot$.
		\item $A^{b} = A(b)$ for every propositional variable $A$.
		\item $(\varphi\circ\psi)^b = \varphi^b\circ\psi^b$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^b = \forall w(b\preceq w\to\varphi^{w}\to\psi^{w})$ where $w$ is some new bound variable.
	\end{itemize}
	Let $K(\varphi)$ encode the theory of Kripke structures, i.e.
	\vspace*{-.1cm}
	\[K(\varphi) = \text{\normalfont PartialOrder}(\preceq)\wedge\forall u\forall w(u\preceq w\to \text{\normalfont Persistent}(u, w))\]
	\vspace*{-.4cm}
	with e.g.
	\vspace*{-.2cm}
	\begin{align*}
		\text{\normalfont PartialOrder}(\preceq) =&\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{\normalfont Persistent}(u, w)=&\:\bigwedge \{A(u)\to A(w)\:|\: \text{ $A$ occurs in $\varphi$}\}
	\end{align*}
	Then define
	$\varphi^{C} = K(\varphi)\to \varphi^{b}$
	where $b$ is a new free variable.
\end{definition}

@@

This is somewhat unsatisfying as we end up in a much more complex logic. However using the normal form from Lemma~\ref{lemma:propositional-normal-form}, i.e. assuming that $\varphi$ is of the form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\right)\to P$$ for some flat clauses $\mathcal R$ and implicational clauses $\mathcal X$, further simplification is possible:
\begin{itemize}
	\item For flat clauses $(\bigwedge A_i\to\bigvee B_j)^b$ is
	$$\forall w\left(b\preceq w\to\bigwedge_i A_i(w)\to\bigvee_jB_j(w)\right).$$
	\item For implication clauses $((A\to B)\to C)^b$ is
	$$\forall w(b\preceq w\to(\forall k(w\preceq k\to A(k)\to B(k)))\to C(w))$$
\end{itemize}

@@

\begin{lemma}
	Suppose $\mathcal M = (M, T)$ is a counter-model to $\varphi^{CH}$, then so is $\mathcal M_T = (M_T, I_T)$.
\end{lemma}

A proof can be found in Appendix~\ref{proof:prop-simplification}.

@@

\begin{corollary}[Small model property]
	If a formula in normal form as in Lemma~\ref{lemma:propositional-normal-form} is intuitionistically invalid then there exists a Kripke counter-model which is a rooted tree of height and degree bounded by $|\mathcal X|$.
\end{corollary}

@@

It is known that a polynomial time translation between QBF and IPC must exist, since both problems are PSPACE complete~\cite{garey1979computers,statman1979intuitionistic}, but we make it explicit. In particular, our translation enables a direct translation of counter-models. For lack of space, we do not give the semantics of QBF in this paper and instead refer the reader to standard sources e.g.~\cite{series/faia/2009-185}.
The translation from the last section will serve as starting point for this section.

@@

\begin{definition}\label{def:fo-translation}
	Let $\varphi$ be a predicate formula. Define $\varphi^{b}$ inductively:
\todo{genau genommen müsstest du noch sagen, dass $b$ eine freie Variable ist. Im letzten Fall der Definition benützen wir dann aber w gleichzeitig als freie und gebundene Variable}
	\begin{itemize}
		\item $\bot^b = \bot$.
		\item $A(\vec t)^{b} = A(\vec t, b)$ for every predicate symbol $A$.
		\item $(\varphi\circ\psi)^b = \varphi^b\circ\psi^u$ for $\circ\in\{\wedge, \vee\}$.
		\item $(\varphi\to \psi)^b = \forall w(b\preceq w\to\varphi^{w}\to\psi^{w})$.
		\item $(\exists x\varphi)^b = \exists x(E(x,b)\wedge\varphi^b)$.
		\item $(\forall x\varphi)^b = \forall w(i\preceq w\to \forall x(E(x, w)\to \varphi^w))$ where $w$ is a new bound variable.
	\end{itemize}
	where $w$ is some new bound variable. Let $K(\varphi)$ encode the theory of Kripke structures as well as the predicate $E$, i.e.
	\begin{align*}
		K(\varphi) = \:& \text{PartialOrder}(\preceq) \wedge \forall u \forall w (u\preceq w\to \text{DomainSubset}(u, w)) \wedge\\
		& \forall u(\text{World(u)}\to \text{DomainClosed}(u))\wedge \forall u\forall w (u\preceq w\to \text{Persistent}(u, w)).
	\end{align*}
	with e.g.:\vspace*{-.76cm}
	\begin{align*}
		\text{PartialOrder}(\preceq) = &\:\forall u(u\preceq u)\wedge\forall u\forall w(u\preceq w\to w\preceq u\to u = w)\wedge\\&\:\forall u\forall v\forall w(u\preceq v\to v\preceq w\to u\preceq w)\\
		\text{DomainSubset}(u, w) = &\:\forall x(E(x, u)\to E(x, w))\\
		\text{World}(u) = &\:\exists xE(x, u)\\
		\text{DomainClosed}(u) = &\:\bigwedge\left\{\forall\vec x(\vec E(\vec x, u)\to E(f(\vec x), u))\:|\:\text{$f$ a function in $\varphi$}\right\}\\
		\text{Persistent}(u, w) = &\:\bigwedge\left\{\forall\vec x(\vec A(\vec x, u)\to A(\vec x, w))\:|\:\text{$A$ a predicate in $\varphi$}\right\}
	\end{align*}
	Then define
	$\varphi^C = K(\varphi)\to\text{World}(b)\to \vec E(\vec a, b)\to \varphi^b$ where $\vec a$ contains all free variables occurring in $\varphi$ and $b$ is a new free variable.
\end{definition}

@@

\begin{proof}
	As in~\cite{otten2005clausal} we may assume that $\varphi$ is of the form $\psi\to P$. Recall that the transformation works by transforming $\psi$ into a set of equivalent assumptions $\left\lbrack \psi\right\rbrack$ of the desired form. We extend the given procedure with rules for quantifiers as follows:
	\begin{align*}
		\left\lbrack(\forall x \psi)\to A\right\rbrack&:= \left\lbrack\psi[s(\vec z)/x]\to A\right\rbrack\\
		\left\lbrack A\to (\forall x\psi)\right\rbrack&:= \{\forall x\chi\:|\:\chi\in\left\lbrack A\to\psi\right\rbrack\}\\
		\left\lbrack(\exists x\psi)\to A\right\rbrack&:= \{\forall x\chi\:|\:\chi\in\left\lbrack A\to\psi\right\rbrack\}\\
		A\to (\exists x\psi)&:= \left\lbrack A\to\psi[s(\vec z)/x]\right\rbrack
	\end{align*}
	where $s$ is a new function symbol, $a$ is a new free variable and $\vec z$ ranges over the free variables in $\psi$.
	While the generated set of clauses is in general not equivalent to $\psi$, section~\ref{section:herbrandiaztion} guarantees that it is at least equivalid.
\end{proof}

\begin{lemma}\label{proof:prop-simplification}
	Suppose $\mathcal M=(M, T)$ is a counter-model to $\varphi^{CH}$, then so is $\mathcal M_T = (M_T, I_T)$.
\end{lemma}
\begin{proof}
	Let $v$ be a variable assignment. It follows directly from the definitions that $\mathcal M_T, v\not\models P(s)$ and $\mathcal M_T, v\models K(\varphi)$. It is easy to show that $\mathcal M_T, v\models \psi$ for $\psi\in\mathcal R^{bS}$ from persistency. Let $\psi\in\mathcal X^{bS}$, we show $\mathcal M_T, v[\psi_1\dots\psi_n/w] \models s\preceq w\to (w\preceq f_\psi(w)\to A(f_\psi(w)\to B(f_\psi(w)))\to C(w))$ for all $\psi_1\dots\psi_n\in M_T$. Let $\psi_1\dots\psi_n\in M_T$ and $m = g_{\psi_n}(\dots(g_{\psi_1}(s^I))$. If $\mathcal M, v[m/w]\models C(w)$ we are done. Otherwise $\mathcal M, v[m/w]\not\models s\preceq w$ or $\mathcal M, v[m/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w)) \to B(f_\psi(w))$. The former is impossible since $g_\psi$ is increasing for all $\psi$. For the latter to hold we must have $\mathcal M, v[m/w]\models w\preceq f_\psi(w)$, $\mathcal M, v[m/w]\models A(f_\psi(w))$ and $\mathcal M, v[m/w]\not\models B(f_\psi(w))$. If $\psi$ is fulfilled at $m$ we have $g_\psi(m) = m$, otherwise $g_\psi(m) = f_{[\psi]}^I(m)$. In any case $\mathcal M_T, v[\psi_1\dots\psi_n/w]\models A(f_\psi(w))$ and $\mathcal M_T, v[\psi_1\dots\psi_n/w]\not\models B(f_\psi(w))$. It follows that $\mathcal M_T, v[\psi_1\dots\psi_n/w]\not\models w\preceq f_\psi(w)\to A(f_\psi(w))\to B(f_\psi(w))$ and therefore $\mathcal M_T, v\models \psi$.
\end{proof}
